Legacy Data Confounds Genomics Studies

### 2019-04-01 legacy.tex
update

>>>>

\subsection{Testing the association of quality to genotype}
When conducting a statistical analysis of population genetics data, 
an essential consideration is  to adjust or account for population structure (i.e. the heterogeneity in biallelic frequencies observed in individuals). Examples of models that widely used to account population structure include the Balding-Nichols model (Balding  Nichols, 1995) \luke{add citation}, and the Pritchard- Stephens-Donnelly model (Pritchard et al., 2000) \luke{add citation}. These and several other simlilar models used in GWAS studies can be understood in terms of the following matrix factorization. 
\begin{align}
\mathbf{L }= \mathbf{AH}
\end{align} 
where the $i$'th column ($\boldsymbol{h}^{(i)}$) of the $K \times I$ matrix $\mathbf{H}$ encodes the population structure of the $i$'th individual and the $s$'th row of the $S \times K$ matrix $\mathbf{A}$ determines how that structure is manifested in SNP $s$. When Hardy-Weinberg equilibrium holds, observed genotype can be assumed to be generated by the following Binomial model.
\begin{align} \label{lfa}
y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big) 
\end{align} 
for $s=1\hdots S$ and $i=i,\cdots, I$, where $y_{si} \in \{0,1,2\}$ and $logit(\pi_{si})$ is the $(s,i)$ element of the matrix $\mathbf{L}$ such that  $\pi_{si}$ is the individual-specific allele frequency.

To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Aassociation Test  (GCAT) proposed by ....  \luke{add citation} The GCAT is a reverse regression approach that assumes the following model.
\begin{align}\label{gcat}
 y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
\end{align} 
for $s=1\hdots S$ and $i=i,\cdots, I$ \luke{Plug in the values of S and I} and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$. The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \luke{add citation} and plugged in the model. We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs. The matrix \luke{Describe matrix}

Under the null hypothesis the slope coefficient is zero $\beta_s=0$ and Model \ref{gcat} reduces to Model \ref{lfa}. $\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$. To test the null hypothesis, the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model. The deviance test statistic is approximately chi-square distributed with one degrees of freedom.

We run a total of $S$ regressions, where $S$ is the total number of genomic loci. Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
